Computational methods in landscape ecology

### 2024-05-28 Sections/introduction.qmd
Several minor cosmetic changes and some progress to simulation section

Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
Linking the spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989; @Turner2005].
Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002; @Wu2013].

<!-- MH: Currently using "computational science", "computational ecology", and "computational methods". Streamline terms?--> 
Computational science involves analyzing abstracted core mechanisms of research questions using data and algorithms.
Thus, computational ecology can be defined as computational science that is used to address ecological research questions with focus on complex, adaptive systems and data-driven approaches [@Poisot2019].
Computational ecology in general and simulations specifically are of importance for ecology because ecological data is often context- and scale-dependent and unfeasible to study in reproducible, replicable, and controllable experimental settings [@Petrovskii2012; but see @Wiersma2022]. Because landscape ecology is a cross-disciplinary field including, e.g., social sciences, geography, or ecology and evolution [@Wiens1997; @With2019], the general increase of available data [@Chi2016; @Jaric2020; @Nathan2022; @Wuest2020], or the complex landscape systems [@Newman2019], there is the need for sophisticated computational methods in order to link patterns and processes. Thus, computational methods are one of the most important tools of modern scientific research [@Prlic2012; @Wilson2014].

This paper aims to introduce the latest developments of computation methods in landscape ecology.
However, it is neither a systematic review of the literature nor general introduction to (computational) landscape ecology.
For a more comprehensive introductions to landscape ecology in general, please see Turner and Gardner [-@Turner2015] or Gergel and Turner [-@Gergel2017].

>>>>

Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989; @Turner2005].
Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002; @Wu2013].

Computational science involves analyzing abstracted core mechanisms of research questions using data and algorithms.
Thus, computational ecology can be defined as computational science that is used to address ecological research questions with focus on complex, adaptive systems and data-driven approaches [@Poisot2019].
Computational ecology in general and simulations specifically are of importance for ecology because ecological data is often context- and scale-dependent and unfeasible to study in controllable, reproducible, and replicable experimental settings [@Petrovskii2012; but see @Wiersma2022].
Because landscape ecology is a cross-disciplinary field including, e.g., social sciences, geography, or ecology and evolution [@Wiens1997; @With2019], the general increase of available data [@Chi2016; @Jaric2020; @Nathan2022; @Wuest2020], or the complex landscape systems [@Newman2019], there is the need for sophisticated computational methods in order to link heterogeneity patterns and ecological processes.
Thus, computational methods are one of the most important tools of modern scientific research [@Prlic2012; @Wilson2014].

Here, we aim to introduce the latest developments of computational methods in landscape ecology.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
For a more comprehensive, general introductions to landscape ecology, please see Turner and Gardner [-@Turner2015], or Gergel and Turner [-@Gergel2017].
First, we will introduce recent computational developments of the most important topics of landscape ecology.
This includes, for example, spatial patterns quantification, connectivity and movement, or model simulations.
Second, we will introduce common software tools that implement or are potentially capable of implementing these developments.
However, due to its many advantages, we will focus on open-source software [@vonKrogh2006], and specifically scripting languages, i.e., R, Python, and Julia.
Last, we will discuss potential future needs developments of computational methods in landscape ecology..

### 2024-05-28 Sections/simulations.qmd
Several minor cosmetic changes and some progress to simulation section

# Simulation models

## Neutral landscapes
<!-- Hesselbarth -->

## Process-based simulation
<!-- Schouten, Hesselbarth -->

>>>>

# Simulation models
<!-- Hesselbarth -->

Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Pascual2005; @Petrovskii2012].
Thus, they can be seen as experimental systems that allow, unlike to nature, all imaginable manipulations in order to advance theoretical developments or test hypotheses [@Peck2004].
Due to the general temporal and spatial scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006; @Synes2016].
In general, simulations models can be classified using two major divisions, namely as *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004;@Synes2016].
Here, we are going to focus mainly on exploratory models, but include both pattern- and process-based models.

## Pattern based
<!-- Hesselbarth -->

### Neutral landscapes models
Neutral landscapes models simulate landscape patterns without assuming any underlying abiotic or biotic processes.
Thus, they are often used as null hypothesis or baselines for comparisons with real landscape patterns, while controlling certain aspects of the landscape [@Li2004;@Wang2008].
The first neutral models simply assigned land cover types randomly to cells in the landscape and were based on percolation theory.
Further developments included hierarchical models that considered different spatial scales while assigning land cover types.
While these first two classes can only simulate landscape with discrete land cover classes, fractal neutral models can also simulate continuous land cover maps by utilizing e.g., Brownian motion.
Last, more complex neutral models were developed that are based on contagion, autocorrelation, or spatial aggregation concepts to simulate more realistic landscapes [@Wang2008].
In order to make multiple neutral landscape model algorithms available several software libraries are available in prominent programming languages including Python [NLMpy; @Etherington2015] or R [NLMR; @Sciaini2018].

### Cellular Automata

## Process based
<!-- Schouten, Hesselbarth -->

[RangeShifter](https://doi.org/10.1111/ecog.05687)

### Landscape Generator
<!-- Hesselbarth -->

[G-RaFFE](https://doi.org/10.1371/journal.pone.0064968)
[LG](https://doi.org/10.1002/ece3.2145)

### Coupled lattice

### IBMs/ABMs

## Misc
[Marine SDMS](https://doi.org/10.1016/j.ecolmodel.2019.108837)
[SDM past present future](https://doi.org/10.1111/jbi.14617)

[Simulations in landscape genetics](https://doi.org/10.1111/j.1365-294X.2010.04678.x)
[HexSim modeling environment](https://doi.org/10.1007/s10980-017-0605-9)
[Digital twins?](https://doi.org/10.1016/j.tree.2023.04.010)
